Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-10 paper.tex
Pass through the vcf2zarr section

\subsection{vcf2zarr}
Converting VCF to Zarr format at Biobank scale is challenging. This main issue is 
around memory usage: although we can view each record in the VCF one-by-one,
we must buffer a full chunk (10,000 variants is reasonable choice) 
in the variants dimension for each of the arrays. 
For VCFs with many FORMAT fields and large numbers of samples this can
easily require tens of gigabytes of RAM per worker, which makes 
parallelism troublesome. Reading the VCF multiple times for different fields
is possible, but would be prohibitively slow for multi-terabyte VCFs.

The \texttt{vcf2zarr} utility solves this problem by first converting 
the VCF data (which can be split across many files) into an Intermediate
Columnar Format (ICF). The \texttt{vcf2zarr explode} command takes a set
of VCFs and stores the data in per-field chunks (by default, 64MiB).
In this way, the memory required is bounded by the size of single record 
BCF record in htslib (plus the size of a decoded bgzf record
[FIXME what is the correct terminology here?]), plus 64MiB for each 
field.  The Intermediate Columnar Format stores the data for each 
field spread across multiple partitions of the data (so that 
multiple parts of the VCF can be read in parallel), and then 
fixed size compressed chunks within those partitions. It also stores information
about the number of records in each partition and chunk of a given
field, so that the record at a particular index can be retrieved efficiently 
later. Summaries such as the minimum and maximum value 
of each field are also maintained, to aid choice of data types later.
A set of VCF files can be converted to intermediate columnar 
format in parallel on a single machine using a given number 
of worker processes using the \texttt{explode} command,
or can be distributed across a cluster using the 
\texttt{dexplode-init},
\texttt{dexplode-partition} and \texttt{dexplode-finalise} commands.

Once the VCF data has converted to the intermediate columnar format,
it can then be converted to Zarr using the \texttt{vcf2zarr encode}
command. Alhough there is a one-to-one mapping between the VCF
fields and the Zarr arrays, there is considerable flexibility 
in how the data is stored, for example in integer width
and compression and filter settings.
By default vcf2zarr chooses integer type sizes based 
on the range of values seen in the intermediate column format
and reasonable compressor defaults, but these can be
altered by generating a schema in JSON format and editing it.
Encoding a given column (for example, \texttt{call\_AD})
involves creating a buffer to hold a full variant-chunk of the 
array in question, and then sequentially filling this buffer with 
values read from the intermediate columnar format. Once the buffer 
is full, the chunks in the samples dimension are compressed
and written to file. [Something about parallelism and distributing]

Fields such as PL are currently problematic when there are large 
numbers of alleles. However, we plan on implementing the 
local alleles strategy~\citep{poterba2024scalable,danecek2021twelve}
which will likely become part of the VCF specification.

>>>>

\subsection{vcf2zarr}
Converting VCF to Zarr at Biobank scale is challenging. 
One problem is to determine the dimension of fields,
(i.e., finding the maximum number of alternate alleles and the 
maximum size of \texttt{Number=.} fields) which requires a full
pass through the data. Another challenge is to keep
memory usage within reasonable limits: although 
we can view each record in the VCF one-by-one,
we must buffer a full chunk (10,000 variants is the default in
\texttt{vcf2zarr}) 
in the variants dimension for each of the fields to convert to Zarr. 
For VCFs with many FORMAT fields and large numbers of samples this can
require tens of gigabytes of RAM per worker, making
parallelism difficult. Reading the VCF multiple times for different fields
is possible, but would be prohibitively slow for multi-terabyte VCFs.

The \texttt{vcf2zarr} utility solves this problem by first converting 
the VCF data (which can be split across many files) into an Intermediate
Columnar Format (ICF). The \texttt{vcf2zarr explode} command takes a set
of VCFs, and reads through them using
\texttt{cyvcf2}~\cite{pedersen2017cyvcf2},
storing each field independently in (approximately) fixed-size 
compressed chunks.
Large files can be partitioned based on information extracted from the 
CSI or Tabix indexes, and so different parts of a file can be 
converted to ICF in parallel.
Once all partitions have completed, information about 
the number of records in each partition and chunk of a given
field is stored so that the record at a particular index 
can be efficiently retrieved. 
Summaries such as maximum dimension and the minimum and maximum value 
of each field are also maintained, to aid choice of data types later.
A set of VCF files can be converted to intermediate columnar 
format in parallel on a single machine 
using the \texttt{explode} command,
or can be distributed across a cluster using the 
\texttt{dexplode-init},
\texttt{dexplode-partition} and \texttt{dexplode-finalise} commands.

Once the VCF data has been converted to the intermediate columnar format,
it can then be converted to Zarr using the \texttt{vcf2zarr encode}
command. We take advantage of Zarr's flexibility
in \texttt{vcf2zarr} by choosing integer widths based 
on the range of values seen in the ICF
and reasonable compressor defaults, but these can be
modified by generating and editing a JSON-formatted storage schema.
Encoding a given column (for example, \texttt{call\_AD})
involves creating a buffer to hold a full variant-chunk of the 
array in question, and then sequentially filling this buffer with 
values read from ICF and flushing to file. 
Similar to the \texttt{explode} command,
encoding to Zarr can be done in parallel on a single 
machine using the \texttt{encode} command,
or can be distributed across a cluster using the 
\texttt{dencode-init},
\texttt{dencode-partition} and \texttt{dencode-finalise} commands.
The distributed commands are fault-tolerant, reporting any failed
parititions so that they can be retried.
